Mapping Mixtures of Genotype-by-Weather Interactions in Switchgrass (Panicum virgatum)

### 2024-01-08 manuscript/revision2/revision2.tex
manuscript: revision 2 text work Abstract through Discussion

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Switchgrass Phenology and Genotype-by-Environment Mapping},
  pdfauthor={Alice H. MacQueen; Li Zhang; Samuel A. Smith; Jason Bonnette; Arvid R. Boe; Phillip A. Fay; Felix B. Fritschi; David B. Lowry; Robert B. Mitchell; Francis M. Rouquette Jr; Yanqi Wu; Arbel Harpak; Thomas E. Juenger},
  pdfkeywords={allele-by-environment effect variation, antagonistic
pleiotropy, photoperiod, cumulative rainfall, genetic variation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title{Switchgrass Phenology and Genotype-by-Environment Mapping}

@@

\significancestatement{The timing of plant seasonal development
(phenology) has major impacts on fitness because of the steep price of
plant-environment mismatches. We infer the mixture of ways genetic
effects on phenological traits depend on plant environment, particularly
how effects covary with weather prior to the phenological event
(GxWeather). Most effects do not have GxWeather, but a minority do.
GxWeather is population-specific. The majority of effects on the start
of vegetative growth in the Gulf population have GxWeather: these
effects differ in sign across environments and covary with a photoperiod
cue 14 days prior. A minority of effects on flowering date in the Gulf
and Midwest populations had GxWeather, covarying with a cumulative
rainfall and a photoperiod cue, respectively.}

@@

\begin{abstract}
The timing of vegetative growth and flowering in plants (``phenological
timings'') depend on both genetic variation and environmental cues;
thus, phenological traits can have important genotype-by-environment
(GxE) interactions. In particular, understanding the mixture and
prevalence of genotype-by-weather (GxWeather) interactions primed by
weather prior to the phenological event should aid prediction and
manipulation of phenological timings. Here, we map genotype-by-weather
effects on phenological timings in two highly divergent switchgrass
(\emph{Panicum virgatum}) populations using repeated plantings of these
populations at sites spanning the central United States. We distinguish
GxWeather patterns covarying with interpretable weather-based cues from
agnostic, site-based patterns. Most GxWeather effects belong to the
latter category and do not covary with weather-based cues. However, 65\%
of Gulf population effects on the start of vegetative growth covary with
daylength 14 days prior to green-up date. 33\% of Gulf population
effects on flowering date covary with cumulative rainfall in the seven
days prior to flowering, while 22\% of Midwest population effects on
flowering date covaried with day length change two days prior to
flowering. An independent pseudo-F2 cross of Gulf and Midwest
individuals mapped 23 additive QTLs for flowering at the same common
gardens, all with significant mash associations and eleven with
enrichment of highly significant mash associations. We demonstrate that
we can identify QTL with GxWeather and assign them to specific
weather-based cues or other patterns. Breeding for particular alleles at
these loci could change flowering responsiveness to photoperiod cues in
switchgrass. More broadly, this approach could be used to identify
genetic marker-environment interactions in any species with related
populations phenotyped in multiple environments.
\end{abstract}

@@

\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\firstpage[10]{2}

The timing of plant vegetative and reproductive development are major
components of plant fitness affected by multiple external environmental
cues (e.g.~degree of winter chilling, day length, temperature, and water
availability) that signal existing or upcoming growing conditions
(1--3). Genetic responses to environmental cues determine the speed,
timing, and energy apportioned to vegetative and reproductive growth and
shape both the individual's lifespan and its lifetime production of
viable seed. Day length (or photoperiod) is one of the most predictable
environmental cues, and genetic sensitivity to photoperiod protects
plants from potentially fatal consequences of phenological responses to
temperature cues at the ``wrong'' time of year. However, the usefulness
of specific environmental cues depends on both features of the
environment and the species' adaptive strategies (4). Species with wide
natural distributions can have multiple distinct environmentally-cued
phenological responses: for example, populations of sunflower
(Helianthus annuus) exhibit day-neutral, facultative short day, and
facultative long-day flowering responses, which vary with their
environments (5, 6). Distinct genetic responses in different
environments are known as genotype by environment interactions, or GxE.

Flowering time, in particular, is a common subject of GxE research
(5--11), a key output of selection driving adaptation to local
environments (2, 12, 13), and a selection target for crop improvement to
adapt crops to local or future environments (14). Changing flowering
responsiveness to photoperiod cues has allowed geographic range
expansion and increased yields in a number of cereal species (15--19)
and other crops (20, 21). Recent statistical advances in studying
phenological GxE have involved determining critical environmental
indices before the phenological event occurs, such as photothermal time
within a critical growth window (10). However, most studies of flowering
GxE focus on finding a single, best fitting form of genotype-environment
covariance, despite the key expectation that different genetic
subpopulations, and even different genomic regions, have likely evolved
distinct patterns of GxE. Additionally, despite theoretical predictions
that local adaptation should involve antagonistic pleiotropy, or
sign-changing GxE, at the level of individual loci (22--25), previous
work has found limited evidence of antagonistic pleiotropy (12, 26), but
has been limited by a known statistical bias that reduced detection of
genetic effects that differ in sign (26--28). Thus, despite substantial
interest in the frequencies of various forms of GxE, the prevalence of
antagonistic pleiotropy relative to other forms of GxE remains unknown.

Switchgrass (\emph{Panicum virgatum}) is considered a short-day plant
with reproductive development strongly linked to day of the year (29).
However, as part of its wide environmental adaptation across the eastern
half of North America, its photoperiodicity has been predicted to differ
by plant latitude of origin (30, 31). We previously found divergent
Midwest and Gulf genetic subpopulations of switchgrass which segregate
for distinct sets of environmental adaptations, based on the analysis of
two measures of fitness (i.e.,biomass and survival) among 732 genotypes
(32). The Midwest genetic subpopulation is primarily composed of
individuals from the well-studied upland switchgrass ecotype (33, 34),
while the Gulf subpopulation has individuals from the lowland ecotype
and the phenotypically intermediate coastal ecotype (32).

Here, we test how the Midwest and Gulf subpopulations differ in their
phenological adaptations across environments, and hence their
phenological GxE. We phenotype a diversity panel of hundreds of
switchgrass genotypes from the Midwest and Gulf subpopulations for the
timing of vegetative development (``green-up'') and reproductive
development (flowering) at eight common garden locations spanning 17
degrees of latitude. These gardens cover the majority of the latitudinal
and climatic range of switchgrass and capture the most comprehensive
picture to date of the environmental variation this species encounters.
After single-site estimation of SNP effects for these phenotypes, we use
multivariate adaptive shrinkage (mash) to jointly estimate SNP effects
in each subpopulation and phenotype at all eight sites (35). Mash allows
us to identify and specify multiple ways genetic effects linked to SNPs
may covary with the environment, and does not have a statistical bias in
detecting SNP effects with the same or opposite signs (36). To confirm
our genetic mapping of GxE, we compare to mapping results from an
outbred pseudo-F2 cross grown at the same sites. Taken together, our
results allow us to describe the environmental cues and genetic
variation affecting phenology in two divergent natural populations of
switchgrass.

@@

In our diversity panel of tetraploid switchgrass (32), genotypes from
the Gulf and Midwest genetic subpopulations had distinct phenological
timings and distinct patterns of phenological correlations across our
eight common garden sites (Figure~\ref{fig-map}). At the three Texas
common gardens (hereafter `Texas' gardens), located within the natural
range of the Gulf subpopulation, Gulf green-up occurred before
Midwestern green-up, and Gulf flowering occurred after Midwestern
flowering (Figure~\ref{fig-map} A). At the four northernmost common
gardens (hereafter `North' gardens), located within the natural range of
the Midwest subpopulation, both Gulf green-up and flowering occurred
after Midwest green-up and flowering. At the Oklahoma common garden,
located near the natural range limits of both the Gulf and the Midwest
subpopulations, Gulf and Midwest green-up occurred over the same time
period. These patterns led to strong negative phenotypic correlations
for green-up between the North and Texas gardens, particularly in the
Gulf and Both subpopulations, and contributed to positive phenotypic
correlations for flowering time of larger magnitude at more northern
gardens (Figure~\ref{fig-map} B).

Narrow-sense heritabilities (h\textsuperscript{2}) indicated that
rank-changing GxE for these phenotypes was present across the common
gardens (Fig 1C). h\textsuperscript{2} were typically high at individual
gardens: 59\% on average for green-up date, and 87\% for flowering date.
However, h2 were variable across gardens, and green-up dates were
uncorrelated (r2 \textless{} 0.2) or negatively correlated between pairs
of gardens (Fig. 1B). These negative and small correlations undoubtedly
contributed to the low h2 values for green-up and flowering date when
estimated jointly at all eight gardens: h2 was 0.8\% for green-up and
23.2\% for flowering date.

We next explored how genetic variation in phenology covaried with
environmental cues by using mash to jointly discover trait-associated
SNPs across all common gardens for the Midwest, Gulf, and `Both'
subpopulations (Fig 1D, SI Appendix, Datasets 1-6). We first estimated
SNP effects using GWAS on site-specific BLUPs for each phenotype. We
then used mash on two subsets of these SNP effects, first to determine
the covariance structures that significantly improved the mash model log
likelihood when included, then to re-estimate the set of SNP effects
with the largest effect estimates in univariate GWAS (hereafter
``strong'' effects) using the set of significant covariance matrices.

Three types of covariance matrices were tested for inclusion:
``canonical'' covariance matrices, with simple patterns of effects;
``data-driven'' matrices derived from common patterns of SNP effects in
the data, and ``hypothesis-based'' covariance matrices that captured the
covariance of weather-based phenological cues for cloned genotypes grown
at multiple common gardens (Table 1, SI Appendix, Section S2). Mash
first assigns mixture proportions for each SNP onto each provided
covariance matrix using maximum likelihood. Then, in a second step, each
SNP's effect estimate undergoes shrinkage such that it is more aligned
with the covariance structure inferred in the first step. The
hypothesis-based covariance matrices are an important advantage mash
offers for studying patterns of GxE: they allow hypothesis testing of
specific environmental drivers for each SNP effect, and loadings on
these matrices provide information on genome-wide patterns of
SNP-environment interaction. Say that a SNP in FLC controls flowering in
a temperature-dependent manner. In that case, that SNP effect could have
a high mixture proportion, or mass, on a covariance matrix created using
a temperature-based environmental cue, such as average temperature on
the day prior to flowering. In our data, we would infer that the effect
of that SNP on flowering was caused by a response to that or a
correlated environmental cue.

The phenotypic correlations for green up date had moderate negative
correlations between the Texas and North gardens, particularly in the
Gulf and in Both subpopulations (Fig 1B). If these phenotypic
correlations have a genetic component, then they might be controlled by
SNP effects that differed in sign across these regions, SNP effects that
are large in one garden or region and non-significant in others, or a
combination of these patterns. Five hypothesis-based covariance matrices
were selected by the greedy mash algorithm, one to two per
subpopulation. Of these five, three had mass on them in mash models of
the strong effects (Fig 2A, 2B). Two of these three matrices had
negative covariances between sites between Texas and North gardens (Fig
2A), while one had all positive or near-zero covariances. SNP-associated
phenotypic effects covaried with different weather-based cues in the
Gulf \& in Both subpopulations (Fig 2B). In total, 65\% of the posterior
weight of strong SNP effects in the mash model of Gulf green-up fell on
a covariance matrix of daylength 14 days prior to the date of green-up.
The covariance matrix for this weather cue was very similar to the
pattern of phenotypic correlation for green-up in the Gulf subpopulation
(Fig 1B; Fig 2A). Mash models of Midwest and Both subpopulation green-up
did not include this covariance matrix; the Midwest had no weight on any
hypothesis-based matrix, while Both subpopulations had non-zero weights
on two additional hypothesis-based matrices, average temperature one day
prior to green-up, and the day length change in seconds in the day prior
to green-up (Fig 2B). The average temperature covariance matrix had
negative covariances between Texas and North gardens, though not as
strong as the negative phenotypic correlations seen in Both
subpopulations (Fig 1B; Fig 2A). Only the Gulf subpopulation and Both
subpopulations had mass on any hypothesis-based covariance matrices (Fig
2C).

For flowering date, distinct weather-based cues captured SNP-associated
effect patterns in the Gulf and Midwest subpopulations. 33\% of SNP
effects on flowering in the Gulf subpopulation covaried with cumulative
rainfall in the seven days prior to flowering (Fig 2D). 22.6\% of SNP
effects on flowering in the Midwest subpopulation contrived with day
length change in the two days prior to flowering (Fig 2D), which had
negative covariances between Texas and North gardens. Neither covariance
matrix was selected for in Both subpopulations, and no hypothesis-based
covariance matrices had mass in Both subpopulations (Fig 2E). In five of
the six mash models of strong effects, the hypothesis-based covariance
matrices captured a minority of the posterior weights of the strong
effects (Figure 2C,E); the majority of this mass was on various
canonical covariance matrices. These matrices included simple
heterozygosity, with intermediate, positive covariances between all
gardens, and garden-specific effects present only at one garden.

@@

We sought additional experimental support for our mash intervals using
an independent pseudo-F2 mapping population created from Gulf \& Midwest
individuals and grown at the same sites (Fig. 4A,B). We conducted
quantitative trait loci (QTL) mapping of flowering as functions of four
environmental cues that we also used as covariance matrices in mash, and
identified eight QTL for flowering date, six QTL for flowering GDD, ten
QTL for flowering day length, and eight QTL flowering day length change,
all of which showed QTL by environment interactions (SI Appendix, Fig.
S3). All QTL for flowering overlapped one or more homologs from rice or
A. thaliana with functionally validated roles in flowering (SI Appendix,
Dataset 7). All flowering QTL intervals contained at least one SNP
significant in at least one mash run at a log10-transformed Bayes Factor
\textgreater{} 2, or in the 1\% tail of significance, whichever was
stricter (SI Appendix, Dataset 8). We also looked for enrichments of
mash SNPs in the 1\% tail of significance (the `mash 1\% tail') within
each QTL interval. At the 5\% level, five QTL had enrichments of SNPs in
the mash 1\% tail. Overall, there were eleven significant enrichments (p
\textless{} 0.05, hypergeometric test) of SNPs in the mash 1\% tail in
the QTL intervals. Our QTL intervals had more enrichments of SNPs in the
mash 1\% tail than were found for all but eleven of these sets of random
genomic intervals (Fig. 4C, p = 0.011). Thus, we were able to
experimentally support our mash intervals with a QTL mapping experiment
using a separate mapping population.

@@

As the climate and the natural environment change, it is increasingly
critical to understand how patterns of gene-environment and
plant-environment interactions will change in response. To do this, we
must understand the current patterns of trait covariation across
environments, the genetic underpinnings of these patterns, and the cases
where this covariation can be altered. Here, we demonstrate that we can
associate multiple patterns of GxE with specific genomic regions using a
switchgrass diversity panel grown at eight common gardens, and also that
we can assign specific SNP-associated patterns of GxE to both
weather-based cues and to other, data-driven patterns. We use this
approach to study GxE in both green-up and flowering phenological data
in the deeply genetically diverged Gulf and Midwest populations of
switchgrass.

Our analysis of green-up in the Gulf and in Both subpopulations revealed
substantial antagonistic pleiotropy in effects between the Texas and
North gardens (Figure 3A). This result supports theoretical models that
local adaptation should involve antagonistic pleiotropy at the level of
individual loci (22--25), and is the first experimental work using QTL
mapping and GWAS across common gardens to find antagonistic pleiotropy
to be common in small genomic regions (12, 36, 37).

Our analysis of flowering showed that the Gulf and Midwest
subpopulations have two distinct photoperiod-related flowering
responses: the Midwest subpopulation is day neutral, and flowering is
cued primarily by a cumulative GDD threshold; in contrast, the Gulf
subpopulation is photoperiod sensitive, and flowering is cued by the
transition to shortening days. This result was supported by observations
that expressing flowering date as a function of the day length at
flowering increased its heritability in the Gulf subpopulation, while
expressing flowering date as a function of cumulative GDD between
green-up and flowering increased the heritability of flowering in the
Midwest subpopulation (Fig. 1D). The genomic regions affecting flowering
found in mash were also supported by QTL from an independent mapping
population (Fig. 4C).

@@

\caption{\label{fig-covar}Figure 2.}

@@

\subsubsection{Mapping major patterns of genotype-by-environment effects
on green-up and
flowering}\label{mapping-major-patterns-of-genotype-by-environment-effects-on-green-up-and-flowering-1}

@@

\subsubsection{Confirmation of genotype-by-environment effects using an
independent mapping
population}\label{confirmation-of-genotype-by-environment-effects-using-an-independent-mapping-population-1}

@@

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Mapping Mixtures of Genotype-by-Weather Interactions in Switchgrass (Panicum virgatum)},
  pdfauthor={Alice H. MacQueen; Li Zhang; Samuel A. Smith; Jason Bonnette; Arvid R. Boe; Phillip A. Fay; Felix B. Fritschi; David B. Lowry; Robert B. Mitchell; Francis M. Rouquette Jr; Yanqi Wu; Arbel Harpak; Thomas E. Juenger},
  pdfkeywords={allele-by-environment effect variation, antagonistic
pleiotropy, photoperiod, cumulative rainfall, genetic variation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title{Mapping Mixtures of Genotype-by-Weather Interactions in
Switchgrass (Panicum virgatum)}

@@

\significancestatement{The timing of plant seasonal development
(phenology) has major impacts on fitness because of the steep fitness
cost of plant-environment mismatches. We infer the mixture of ways
genetic effects on phenological traits depend on plant environment,
focusing on how effects covary with weather prior to the phenological
event (GxWeather). Most effects do not have GxWeather, but a minority
do. GxWeather is population-specific. The majority of effects on the
timing of vegetative growth in the Gulf population have GxWeather: these
effects differ in sign across environments and covary with a photoperiod
cue two weeks prior. A minority of effects on flowering date in the Gulf
and Midwest populations had GxWeather, covarying with a cumulative
rainfall and a photoperiod cue, respectively.}

@@

\begin{abstract}
The timing of vegetative and reproductive growth in plants
(``phenological timings'') depends on both genetic variation and
environmental cues; thus, phenological traits can have important
genotype-by-environment (GxE) interactions. Identifying the mixture and
prevalence of GxE primed by weather prior to the phenological event
(GxWeather) should aid prediction and manipulation of phenological
timings. Here, we map GxWeather effects on phenological timings in two
highly divergent switchgrass (\emph{Panicum virgatum}) populations using
repeated plantings of these populations at eight sites spanning the
central United States. We distinguish GxWeather patterns covarying with
interpretable weather-based cues from agnostic, site-based patterns.
Most GxWeather effects belong to the latter category and do not covary
with weather-based cues. However, 65\% of effects on the timing of
vegetative growth in the Gulf population covary with daylength 14 days
prior to green-up date. 33\% of effects on the timing of flowering in
the Gulf population covary with cumulative rainfall in the seven days
prior to flowering, while 22\% of effects on flowering timing in the
Midwest population covary with day length change two days prior to
flowering. An independent pseudo-F2 cross of Gulf and Midwest
individuals at the same sites mapped 23 additive QTLs for flowering, all
with significant associations and eleven with enrichment of highly
significant associations in our model of GxWeather effects. We
demonstrate that we can identify QTL with GxWeather and assign them to
specific weather-based cues or other patterns. Breeding for particular
alleles at these loci could change flowering responsiveness to
photoperiod and rainfall cues in switchgrass. More broadly, this
approach could be used to identify genetic marker-environment
interactions in any species with related populations phenotyped in
multiple environments.
\end{abstract}

@@

\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
\firstpage[9]{2}

Plant phenological timings are major components of plant fitness
affected by multiple external environmental cues (e.g.~degree of winter
chilling, day length, temperature, and water availability) that signal
existing or upcoming growing conditions (1--3). Genetic responses to
environmental cues determine the speed, timing, and energy apportioned
to vegetative and reproductive growth and shape both the individual's
lifespan and its lifetime production of viable seed. Day length (or
photoperiod) is one of the most predictable environmental cues, and
genetic sensitivity to photoperiod protects plants from potentially
fatal consequences of phenological responses to temperature cues at the
``wrong'' time of year. However, the usefulness of specific
environmental cues depends on both features of the environment, such as
cue predictability and relevance, and the species' adaptive strategies
(4). Species with wide natural distributions can have multiple distinct
environmentally-cued phenological responses: for example, populations of
sunflower (Helianthus annuus) exhibit day-neutral, facultative short
day, and facultative long-day flowering responses, which vary with their
environments (5, 6). Distinct genetic responses in different
environments are known as genotype by environment interactions, or GxE.

Flowering time, or the transition from vegetative to reproductive
growth, is a common subject of GxE research (5--11), a key output of
selection driving adaptation to local environments (2, 12, 13), and a
selection target for crop improvement to adapt crops to local or future
environments (14). Changing flowering responsiveness to photoperiod cues
has allowed geographic range expansion and increased yields in a number
of cereal species (15--19) and other crops (20, 21). Recent statistical
advances in studying phenological GxE have involved determining critical
environmental indices before the phenological event occurs, such as
photothermal time within a critical growth window (10). However, most
studies of flowering GxE focus on finding a single, best fitting form of
genotype-environment covariance, despite the key expectation that
different genetic subpopulations, and even different genomic regions,
have likely evolved distinct patterns of GxE. Additionally, despite
theoretical predictions that local adaptation should involve
antagonistic pleiotropy, or sign-changing GxE, at the level of
individual loci (22--25), previous work has found limited evidence of
antagonistic pleiotropy (12, 26). However, this work has been limited by
a known statistical bias that reduced detection of genetic effects that
differ in sign (26--28). Thus, despite substantial interest in the
frequencies of various forms of GxE, the prevalence of antagonistic
pleiotropy relative to other forms of GxE remains unknown.

Switchgrass (\emph{Panicum virgatum}) is considered a short-day plant
with reproductive development strongly linked to day of the year (29).
However, as part of its wide environmental adaptation across the eastern
half of North America, its photoperiodicity has been predicted to differ
by plant latitude of origin (30, 31). We previously found divergent
Midwest and Gulf genetic subpopulations of switchgrass with distinct
sets of environmental adaptations, in that both populations had distinct
genetic variation associated with each of two fitness proxies, biomass
and survival (32). The Midwest genetic subpopulation is primarily
composed of individuals from the well-studied upland switchgrass ecotype
(33, 34), while the Gulf subpopulation has individuals from the
well-studied lowland ecotype and the phenotypically intermediate coastal
ecotype (32).

Here, we estimate the ways genetic effects on phenological timings
depend on weather prior to the phenological event. To do this, we
phenotype a diversity panel of hundreds of switchgrass genotypes from
the Midwest and Gulf subpopulations for the timing of vegetative and
reproductive development at eight common garden locations spanning 17
degrees of latitude. These gardens cover the majority of the latitudinal
and climatic range of switchgrass and capture the most comprehensive
picture to date of the environmental variation this species encounters.
We define a number of ways phenological timings might covary with
weather (@tbl-covar) and additional ways phenological timings might vary
by site (@SI-info), then jointly re-estimate genetic effects on these
timings at all eight sites using the set of these covariance types that
significantly improved the modeled log-likelihood when included
(@SI-info) (35). The method we use for effect re-estimation, mash,
allows us to identify and specify multiple ways genetic effects may
covary with the environment, and does not have a statistical bias in
detecting genetic effects with the same or opposite signs (36). To
confirm our genetic mapping of GxWeather, we compare our re-estimated
genetic effects to mapping results from an outbred pseudo-F2 cross grown
at the same sites. Our analysis allow us to describe the weather cues
and genetic variation affecting phenology in two divergent natural
populations of switchgrass.

@@

Genotypes from the Gulf and Midwest subpopulations had distinct
phenological timings and distinct patterns of phenological correlations
across our eight common garden sites (Figure~\ref{fig-map}). At the
three Texas common gardens (hereafter `Texas' gardens), located within
the natural range of the Gulf subpopulation, Gulf vegetative growth
occurred before Midwestern vegetative growth, and Gulf flowering
occurred after Midwestern flowering (Figure~\ref{fig-map} A). At the
four northernmost common gardens (hereafter `North' gardens), located
within the natural range of the Midwest subpopulation, both Gulf
vegetative growth and flowering occurred after Midwest vegetative growth
and flowering. At the Oklahoma common garden, located near the natural
range limits of both the Gulf and the Midwest subpopulations, Gulf and
Midwest vegetative growth occurred over the same time period, and Gulf
flowering occurred after Midwestern flowering (Figure~\ref{fig-map} A).
These patterns led to strong negative phenotypic correlations for the
timing of vegetative growth between the North and Texas gardens,
particularly in the Gulf and across all individuals (hereafter, `Both'
subpopulations), and contributed to positive phenotypic correlations for
the timing of flowering that had larger magnitudes at more northern
gardens (Figure~\ref{fig-map} B).

Narrow-sense heritabilities (h\textsuperscript{2}) indicated that
rank-changing GxE for these phenotypes was present across the common
gardens (Figure~\ref{fig-map} C). h\textsuperscript{2} were typically
high at individual gardens: 59\% on average for green-up date, and 87\%
for flowering date. However, h\textsuperscript{2} were variable across
gardens, and green-up dates were uncorrelated (r\textsuperscript{2}
\textless{} 0.2) or negatively correlated between pairs of gardens
(Figure~\ref{fig-map} B). These negative and small correlations
undoubtedly contributed to the low h\textsuperscript{2} values for
green-up and flowering date when estimated jointly at all eight gardens:
h\textsuperscript{2} was 0.8\% for green-up and 23.2\% for flowering
date.

\subsubsection{GxWeather mixtures for vegetative and reproductive
timing}\label{gxweather-mixtures-for-vegetative-and-reproductive-timing}

We next looked for evidence of GxWeather by jointly re-estimating
genetic effects across all eight common gardens using a model that
allows effects to covary between gardens in many ways
(Figure~\ref{fig-map} D, SI Appendix, Datasets 1-6). To do this, we used
genome-wide association (GWAS) on site-specific BLUPs for the timing of
vegetative growth and flowering to determine preliminary single
nucleotide polymorphism (SNP) effect estimates at each garden. We then
jointly modeled genetic effect estimates, and the mixture of ways these
effects might covary, across all eight common gardens, using a two-part
procedure. We first modeled genetic effects with different combinations
of covariance types, using a subset of effects selected at random from
all SNPs used in the GWAS. In these models, we specified many ways
genetic effects could covary between gardens (as expanded on below) and
used a greedy algorithm to iteratively select covariance structures that
significantly improved the log likelihood of a model when included.
Given the high correlation between some of the covariance patterns we
tested, this approach allowed us to select only the subset of patterns
of covariation that most improved the fit of the joint effect patterns
present in the data. We then re-estimated the set of genetic effects
that had the largest effect estimates in univariate GWAS (hereafter
``strong'' effects) using the set of significant covariance matrices
that most improved the model fit on the random effects.

We created three categories of ways effects could covary across gardens:
``canonical'' covariance, with simple patterns of effects;
``data-driven'' covariances derived from common patterns of SNP effects
in the data, and ``GxWeather'' covariances that captured the covariance
of weather variables at specific times before the phenological event
(Table 1, SI Appendix, Section S2). For example, GxWeather covariance
between gardens in MI and MO, for rainfall in the seven days prior to
flowering, captures the covariance in the amount of rainfall genotype A,
B, and C receive in the seven days prior to their three flowering dates
in MI and their three flowering dates in MO. If these six rainfall
values are similar, covariance will be near one; if they are
uncorrelated, covariance will be near zero; if there is sufficient rank
changing in rainfall amounts between gardens for these genotypes, this
covariance will be near negative one.

The GxWeather covariances allow hypothesis testing of specific weather
variables as cues for each genetic effect. Say that a SNP in FLC
controls flowering in a photoperiod-dependent manner. In that case, that
SNP effect could have a high mixture proportion, or mass, on a
covariance matrix created using a photoperiod-based environmental cue,
such as day length two days prior to flowering. In our data, we would
infer that the effect of that SNP on flowering was caused by a response
to that or a correlated environmental cue. Overall, loadings of genetic
effects on these GxWeather matrices in the model with re-estimated
effects provide information on genome-wide patterns of SNP-environment
interaction.

The phenotypic correlations for the timing of vegetative growth had
moderate negative correlations between the Texas and North gardens,
particularly in the Gulf and in Both subpopulations
(Figure~\ref{fig-map} B). If these phenotypic correlations have a
genetic component, then they might be controlled by SNP effects that
differed in sign across these regions (effects with antagonistic
pleiotropy), SNP effects that are large in one garden or region and
non-significant in others (effects with amplification), or a combination
of these patterns. Five GxWeather covariance matrices were selected by
the greedy mash algorithm, one to two per subpopulation. Of these five,
three had mass on them in mash models of the strong effects
(Figure~\ref{fig-covar} A-B). Two of these three matrices had negative
covariances between sites between Texas and North gardens
(Figure~\ref{fig-covar} A), while one had all positive or near-zero
covariances. SNP-associated phenotypic effects covaried with different
weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the mash model of Gulf vegetative growth fell on a
covariance matrix of daylength 14 days prior to the date of vegetative
growth. The covariance matrix for this weather cue was very similar to
the pattern of phenotypic correlation for the timing of vegetative
growth in the Gulf subpopulation (Figure~\ref{fig-map} B;
Figure~\ref{fig-covar} A). Mash models of the timing of vegetative
growth in Midwest and Both subpopulations did not include this GxWeather
covariance; the Midwest had no weight on any GxWeather covariance, while
Both subpopulations had non-zero weights on two additional GxWeather
covariance types, average temperature one day prior to green-up, and the
day length change in seconds in the day prior to green-up
(Figure~\ref{fig-covar} B). The average temperature covariance matrix
had negative covariances between Texas and North gardens, though not as
strong as the negative phenotypic correlations seen in Both
subpopulations (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Only
the Gulf subpopulation and Both subpopulations had mass on any GxWeather
covariance matrices (Figure~\ref{fig-covar} C).

For flowering date, distinct GxWeather covariances captured
SNP-associated effect patterns in the Gulf and Midwest subpopulations.
33\% of SNP effects on flowering in the Gulf subpopulation covaried with
cumulative rainfall in the seven days prior to flowering (Fig 2D).
22.6\% of SNP effects on flowering in the Midwest subpopulation covaried
with day length change in the two days prior to flowering (Fig 2D),
which had negative covariances between Texas and North gardens. Neither
covariance matrix was selected for in Both subpopulations, and no
GxWeather covariance matrices had mass in Both subpopulations (Fig 2E).
In five of the six mash models of strong effects, the GxWeather
covariance matrices captured a minority of the posterior weights of the
strong effects (Figure 2C,E); the majority of this mass was on various
canonical covariance matrices. These matrices included simple
heterozygosity, with intermediate, positive covariances between all
gardens, and garden-specific effects present only at one garden.

@@

\subsubsection{Confirmation of effects on phenology using an independent
mapping
population}\label{confirmation-of-effects-on-phenology-using-an-independent-mapping-population}

We sought additional experimental support for our significant effect
re-estimates using an independent pseudo-F2 mapping population created
from Gulf \& Midwest individuals and grown at the same sites (Fig.
4A,B). We conducted quantitative trait loci (QTL) mapping of flowering
as functions of four environmental cues that we also used as covariance
matrices in mash, and identified eight QTL for flowering date, six QTL
for flowering GDD, ten QTL for flowering day length, and eight QTL
flowering day length change, all of which showed QTL by environment
interactions (SI Appendix, Fig. S3). All QTL for flowering overlapped
one or more homologs from rice or \emph{A. thaliana} with functionally
validated roles in flowering (SI Appendix, Dataset 7). All flowering QTL
intervals contained at least one SNP significant in at least one mash
run at a log10-transformed Bayes Factor \textgreater{} 2, or in the 1\%
tail of significance, whichever was stricter (SI Appendix, Dataset 8).
We also looked for enrichments of mash SNPs in the 1\% tail of
significance (the `mash 1\% tail') within each QTL interval. At the 5\%
level, five QTL had enrichments of SNPs in the mash 1\% tail. Overall,
there were eleven significant enrichments (p \textless{} 0.05,
hypergeometric test) of SNPs in the mash 1\% tail in the QTL intervals.
Our QTL intervals had more enrichments of SNPs in the mash 1\% tail than
were found for all but eleven of these sets of random genomic intervals
(Fig. 4C, p = 0.011). Thus, we were able to experimentally support our
significant effect re-estimates with a QTL mapping experiment using a
separate mapping population.

@@

As the climate and the natural environment change, it is increasingly
critical to understand how patterns of gene-environment and
plant-environment interactions will change in response. To do this, we
must understand the current patterns of trait covariation across
environments, the genetic underpinnings of these patterns, and the cases
where this covariation can be altered. Here, we demonstrate that we can
associate multiple patterns of GxWeather with specific genomic regions
using a switchgrass diversity panel grown at eight common gardens. We
can assign genetic effects to both GxWeather patterns with interpretable
weather-based cues, and to agnostic, site-based patterns. We use this
approach to study GxWeather for the timings of vegetative and
reproductive development in the deeply genetically diverged Gulf and
Midwest subpopulations of switchgrass.

Our analysis of the timing of vegetative growth in the Gulf and in Both
subpopulations revealed substantial antagonistic pleiotropy in effects
between the Texas and North gardens (Figure 3A). This result supports
theoretical models that local adaptation should involve antagonistic
pleiotropy at the level of individual loci (22--25), and is the first
experimental work using QTL mapping and GWAS across common gardens to
find antagonistic pleiotropy to be common in small genomic regions (12,
37, 38).

Our analysis of the timing of flowering showed that the Gulf and Midwest
subpopulations have distinct GxWeather: flowering timing in the Midwest
subpopulation has photoperiod-related variation, in that flowering
timing covaries with a day length change signal in the two days before
flowering occurs. In contrast, the Gulf subpopulation does not have
variation in flowering based on photoperiod. Instead, the Gulf
subpopulation has variation in flowering that covaries with the rainfall
that occurs in the week prior to flowering. The genomic regions
affecting flowering that we re-estimated across all eight sites were
also supported by QTL from an independent mapping population at these
sites (Fig. 4C).

@@

\caption{\label{fig-covar}Caption for this figure\ldots{}}

@@

@@

@@

\bibitem[\citeproctext]{ref-savolainen2013ecological}
\CSLLeftMargin{37. }%
\CSLRightInline{O. Savolainen, M. Lascoux, J. Merilä, Ecological
genomics of local adaptation. \emph{Nature Reviews Genetics}
\textbf{14}, 807--820 (2013).}

\bibitem[\citeproctext]{ref-lowry2019qtl}
\CSLLeftMargin{38. }%
\CSLRightInline{D. B. Lowry, \emph{et al.}, QTL\(\times\) environment
interactions underlie adaptive divergence in switchgrass across a large
latitudinal gradient. \emph{Proceedings of the National Academy of
Sciences} \textbf{116}, 12933--12941 (2019).}
